---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file and re-build the .md with devtools::build_readme() -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# casteval

<!-- badges: start -->
<!-- badges: end -->

## Overview

`{casteval}` facilitates the evaluation of time series forecasts.

`r emo::ji("white heavy check mark")` Flexibly accepts forecasts as either a set of individual model realizations or as the summary of an ensemble of realizations via quantiles at each time point

`r emo::ji("white heavy check mark")` Includes multiple scoring methods

`r emo::ji("white heavy check mark")` Provides a suite of visualiation tools to help assess forecast and scoring performance

## Installation

The simplest way to install `{casteval}` is to use the [`remotes`](https://remotes.r-lib.org/) package:

```{r, eval = FALSE}
# install.packages("remotes")
remotes::install_git("https://gitlab.cscscience.ca/phrsd/risk/casteval.git")
```

If you do not wish to install the `remotes` package, or have trouble using it, you can simply clone this repository locally and then use:
```{r, eval = FALSE}
install.packages("your/path/to/casteval", type="source")
```

## Usage

See `vignette("casteval")` to get started.

## For developers

### Source code

The source code is located [here](https://gitlab.cscscience.ca/phrsd/risk/casteval).

### `renv`

This project uses [`renv`](https://rstudio.github.io/renv/index.html) to manage dependencies.

For R package projects, `renv` tries to intelligently discover dependencies by looking at the fields in the DESCRIPTION file that are relevant for users, like `Imports` and `Depends` (see `renv::settings$package.dependency.fields()`). For developers, the `Suggests` field can be equally important to track as it can include packages used in the process of development but not in the actual source code of the package, like `devtools`, `covr`. This project's GitLab CI pipeline relies on both of these packages, and the only reference it has for which packages to install/load when running is from the `renv` lockfile. Thus, it's important for us to track our development packages with `renv`. 

The `renv` FAQ page [recommends tracking development dependencies in `Suggests`](https://rstudio.github.io/renv/articles/faq.html#how-should-i-handle-development-dependencies). It doesn't, however, seem to tell you that in order to use their recommendation, you need to use the `dev = TRUE` flag in `renv::status()` to accurately check the status of such a project, as well as in `renv::snapshot()` when recording files (the default is `dev = FALSE`). If you simply execute `renv::snapshot(dev = FALSE)`, `renv` will miss/remove the development dependencies in `Suggests`, which will break our CI pipelines. Once these dependencies have been recorded, `renv::status(dev = FALSE)` will report that the project is out-of-sync, as it will see installed dependencies that are not used anywhere (since it skips `Suggests`).

The bottom line is to be sure to use the `dev = TRUE` option when working in this repository. If you need to add any development dependencies to the project, be sure to add them to the `Suggests` field of the `DESCRIPTION`, run `renv::install(dev = TRUE)` to install them, then `renv::snapshot(dev = TRUE)` to record them.

A side effect of this default behaviour is that, when you open a new R session in this project, `renv::status()` will get called and report that the project is out-of-sync. Don't trust this message! Run `renv::status(dev = TRUE)` to check the true status of the project. (There is a [feature request](https://github.com/rstudio/renv/issues/1760) to enable a user to configure the default behaviour of `renv::status()` on a project-by-project basis specifically to accommodate the above situation, but this issue is still open as of this writing on 2024-06-11...)

### Test coverage

You can use `covr::report()` to periodically check test coverage locally in development.