---
title: "Evaluate time series forecasts with casteval"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{casteval}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  strip.white=TRUE
)

library(casteval)

ggplot2::theme_set(ggplot2::theme_light())
```

# Overview

`{casteval}` is an R package that helps you automate the evaluation of time series forecasts.
It provides functionality for formatting, processing, scoring, and visualizing forecasts.

A typical workflow using `{casteval}` is:

1. [**Create a forecast object**](#create-a-forecast-object) by inputting a forecast and its metadata into `create_forecast()`
2. [**Score the forecast**](#score-a-forecast) using `score()`, which accepts forecast object(s), corresponding observations, and a particular scoring function
3. [**Visualize the forecast**](#visualize-forecast-evaluation) using `plot_forecast()` with a scored forecast and observations

This vignette guides you through this process. 
<!-- In this vignette, we demonstrate `{casteval}`'s main functionality via deliberately simple examples.
For a more realistic start-to-finish demonstration of the package's usage with real forecasts, please see [`vignette("denmark2020")`](denmark2020.html). -->

# Create a forecast object {#create-a-forecast-object}

In order to create a forecast object, a forecast must first be cast in one of the accepted formats, namely as a [data frame](#forecast-data-frame) or as a [list](#forecast-list).

## Forecast input as data frames {#forecast-data-frame}

**Forecast data frames** must contain:

1. a `time` column, which can contain either numbers, dates, or date-times
1. forecast data, which can be passed as raw or summarized. 

**Raw forecast data** are the individual realizations of a model. **Summarized forecast data** describes an ensemble of realizations by quantiles computed for each time point.

### Raw forecast data

Raw forecast data must be passed in a column named `val`. A `sim` column may be provided optionally to identify different realizations (enables some scoring and plotting features).

```{r fdf_raw}
# raw forecast over with numeric times, without simulation numbers
data.frame(
  time=rep(1:5,each=3),
  val=c(100,101,102,110,111,112,120,121,122,130,131,132,140,141,142)
)

# raw forecast data with dates, with simulation numbers
data.frame(
  time=lubridate::as_date(rep(1:5,each=3)),
  sim=rep(1:3,5),
  val=c(100,101,102,110,111,112,120,121,122,130,131,132,140,141,142)
)
```

### Summarized forecast data

Summarized forecast data may be stored in the following columns:

- Quantile columns must start with `val_q` followed by a number from 0 to 100, _e.g._, `val_50` would be the 50th quantile, _i.e._, the median
- The mean column must be named `val_mean`

Note that summarized data never contains a `sim` column.

```{r fdf_summarized}
# summarized forecast data with the 25th, 50th, and 75th percentiles
data.frame(
  time=lubridate::as_datetime(1:5),
  val_q25=c(100,105,103,104,105),
  val_q50=c(201, 210, 205, 201, 200),
  val_q75=c(304, 305, 303, 303, 302)
)

# summarized forecast data with the mean, and the 2.5th and 97.5th percentiles
data.frame(
  time=1:5,
  val_q2.5=c(100,103,104,105,102),
  val_mean=c(150,155,160,155,154),
  val_97.5=c(200,200,2204,205,206)
)
```

All forecast data columns (`val`, `val_*`) as well as the `sim` column must be numeric.

## Forecast input as lists {#forecast-list}

Forecast data frames are straightforward representations forecasts, though they are innefficient for raw forecast data as the `time` column gets repeated for each realization. If you have particularly long forecasts and/or many realizations, forecast data frames can quickly grow large, which can be slow to work with. Instead, you may wish to represent your forecast as a list.

`{casteval}` accepts raw forecast data as a named list with the following entries:

  - `time`: a single vector of times as numbers, dates, or date-times
  - `vals`: a list of numeric vectors containing the forecasted values, one per realization, where each vector is the same length as `time`

Since the individual realizations can be inferred from this casting, there is no need to provide additional information to identify individual realizations (as we did with the `sim` column [above](#forecast-data-frame)) in order to enable related downstream features.

## `create_forecast()` {#forecast-object}

**Forecast objects** are created with `create_forecast()`.
Its first argument is `dat`, which is the forecast data in one of the formats described [above](#create-a-forecast-object).

`create_forecast()` accepts forecast metadata as additional, optional arguments:

- `name`: a label for the forecast. Plotting functions will use `name` to title the plots they produce.
- `forecast_time`: the time at which the forecast period begins in `dat`^[Time series forecasts can typically be divided into a fit period and a forecast period, distinguished by whether or not observations were available at the time of forecast, respectively. It can be useful in visualizing and/or scoring forecasts to distinguish between these two periods.]. The `forecast_time` format must match that of the times in the `time` entry of `dat`, and can be used in scoring functions to compute scores only after times relative to `forecast_time`.

```{r create_forecast}
# forecast data with 4 time points and an ensemble of 3 simulations
dat1 <- list(
  time=1:4,
  vals=list(
    c(100, 102, 110, 108),
    c(200, 195, 197, 196),
    c(300, 301, 300, 302)
  )
)

fc1 <- create_forecast(dat1, name="forecast 1", forecast_time=3)

# the same data but formatted as a forecast data frame
dat2 <- data.frame(
  time=rep(1:4, each=3),
  sim=rep(1:3, times=4),
  val=c(100, 200, 300, 102, 195, 301, 110, 197, 300, 108, 196, 302)
)

fc2 <- create_forecast(dat2, name="forecast 2", forecast_time=3)

print(fc1)

# aside from the order of rows, the resulting forecast data frames are identical
waldo::compare(dplyr::arrange(fc1$data, time), fc2$data)

# mean-and-quantiles forecast data
dat3 <- data.frame(
  time=1:5,
  val_q2.5=c(100,103,104,105,102),
  val_mean=c(150,155,160,155,154),
  val_q97.5=c(200,200,2204,205,206)
)

# note how we omit `forecast_time`
fc3 <- create_forecast(
  dat3,
  name="forecast 3"
)

print(fc3)
```

`create_forecast()` returns a forecast object, which is a list with the following fields:

- `data`: a [forecast data frame](#forecast-data-frame)
- `name`: the name of the forecast, if provided
- `forecast_time`: when the forecast was made, if provided 

```{r forecast_object}
# a forecast object with no optional metadata
list(
  data=data.frame(
    time=1:5,
    val=6:10
  ),
  name=NULL,
  forecast_time=NULL
)

# a forecast object with metadata
list(
  data=data.frame(
    time=lubridate::as_date(1:5),
    val=6:10
  ),
  name="A forecast",
  forecast_time=lubridate::as_date(3)
)
```

This list could be created by hand, without the help of `create_forecast()`, but we recommend the latter approach as `create_forecast()` validates its inputs to avoid issues downstream.
For example, it checks that:

- `forecast_time`'s type is consistent with `dat`
- quantile values are logically possible (_e.g._, median values can't be smaller than the 25th quantiles)
- there are no conflicting rows (_e.g._, there can't be two rows with the same time reporting different means)

# Score a forecast {#score-a-forecast}

Forecasts are scored against observations, which must be passed to `{casteval}` in a specific format.

## Observations data frames {#observations-data-frame}

Observations are passed to `{casteval}` in a data frame with a `time` column and a `val_obs` column.
Similar to [forecast data frames]{#forecast-data-frame}, `time` may be either numeric, dates, or date-times.
`val_obs` must be numeric.

```{r obsdf}
# An observations data frame
data.frame(
  time=1:5,
  val_obs=c(50,60,55,57,59)
)
```

## Scoring functions

We score a forecast using `score(fcsts, obs, fun, ...)`.

- `fcsts` is either a forecast object or a list of forecast objects
- `obs` is an observations data frame
- `fun` is a scoring function
- `...` are additional parameters which will be passed to `fun`

Scoring functions which can be passed to `score()` include:

- `accuracy()`, which calculates the success rate of a forecast's quantile range predictions
- `crps()`, which calculates the Continuous Ranked Probability Score (CRPS) of a forecast
- `log_score()`, which calculates the (positive) logarithmic score of a forecast
- `bias()`, which calculates how much a forecast overpredicts/underpredicts values

`score()` scores every forecast passed to it against `obs`, and returns a score for each one.^[If `fcsts` is a list of forecast objects, a list of scores will be returned. If `fcsts` is a single forecast object, a single score will be returned]

Below are examples of scoring.
For a detailed explanation of how scoring works, see the [scoring vignette](scoring.html)


# Visualize forecast evaluation {#visualize-forecast-evaluation}

`{casteval}` provides plotting functions which allow you to visualize your forecasts, observations, and sometimes scores.
Several modular functions (`plot_ensemble()`, `plot_observations()`, etc.) implement individual plotting functionality, while the more user-friendly `plot_forecast()` combines them all.
We describe these functions below.

## Plot a forecast

`plot_forecast(fcst, obs=NULL, quant_pairs=NULL, invert_scale=FALSE, score=NULL)` plots a forecast and optionally overlays observations, quantile intervals, and color-codes the observations based on score. 

`quant_pairs` is like the parameter with the same name in `accuracy()`, though here it is only used to draw quantile bands on the plot.

If `val_q50` is present in the forecast data, then the median will be plotted.
Similarly, if `val_mean` is present, the mean will be plotted.

```{r plot_forecast_examples}
# Create a forecast
fc <- create_forecast(list(
  time=1:10,
  vals=list(
    c(1,2,3,5,4,5,4,6,6,5),
    c(1,3,5,4,6,5,7,9,8,8),
    c(1,4,3,4,5,6,5,3,2,2),
    c(1,2,4,5,7,8,7,9,10,9)
  )
  ),
  name = "Forecast date: 2024-07-30"
)

# Plot it
plot_forecast(fc)

# Plot it and display 3 quantile intervals
plot_forecast(fc, quant_pairs=list(c(25,75), c(2.5,97.5), c(5,95)))

# Create a forecast with mean and median data
fc2 <- create_forecast(data.frame(
  time=1:5,
  val_mean=6:10,
  val_q50=c(5.5,6,7.5,8,8)
))

# Plot it
plot_forecast(fc2)
```

### Plot observations with scores

`plot_forecast()` function computes scores internally, so there is no need to precompute them as we did [above](#score-a-forecast). (However, it is important to understand the scoring functions available, in order to use `plot_forecast()`.) Just pass a compatible scoring function (one that supports the `summarize` flag) to the `score` argument. Then, `plot_forecast()` will run `score(fcst, obs, summarize=FALSE)`, and use the resulting data frame's `score` column to color code the observation points.

```{r plot_forecast_examples2}
# Create some observations
obs <- data.frame(time=1:10, val_obs=c(1,4,8,10,11,8,5,3,3,2))

# Plot them over the forecast and its quantile intervals
plot_forecast(fc, obs=obs, quant_pairs=list(c(25,75), c(2.5,97.5), c(5,95)))

# Colour code points using `log_score()`
# `plot_forecast()` automatically passes `summarize=FALSE` to
# scoring functions in order to obtain the score for each day
plot_forecast(fc, obs=obs, score=log_score, quant_pairs = c(2.5, 97.5))
```

If you need to pass additional arguments to a scoring function, it needs to be wrapped so that it only accepts `fcts`, `obs`, and `summarize`. `{casteval}` contains the utilities `make_accuracy()` and `make_log_score()` to easily make wrapped scoring functions. For example, `accuracy()` needs a quantile pair specified, so we can do:

```{r plot_forecast_examples3}
# We use a function factory to wrap accuracy() (see `?make_accuracy`)
quant <- c(5,95)
acc90 <- make_accuracy(quant)

# Identify the observations inside the 5%-95% quantile interval
pp <- plot_forecast(fc, obs=obs, quant_pairs=list(c(25,75), c(2.5,97.5), c(5,95)), score=acc90)
pp
```

The `invert_scale` parameter allows you to invert the color scale for observations.
For example, for Continuous Ranked Probability Score, a lower score is better.

```{r invert_scale}
fc <- create_forecast(list(
  time=1:3,
  vals=list(c(4,7,8), c(5,6,7), c(4,6,6))
))
obs <- data.frame(time=1:3, val_obs=5:7)

# don't invert the scale (default behaviour)
plot_forecast(fc, obs, score=crps, invert_scale=FALSE)

# invert the scale
plot_forecast(fc, obs, score=crps, invert_scale=TRUE)
```

### Customize the output
Since `plot_forecast()` returns a `ggplot2::ggplot` object, you can further customize the plot to suit your application:

```{r plot_forecast_examples4}
(pp
  + ggplot2::labs(
    x = "Days from forecast",
    y = "Hospital admissions",
    colour = glue::glue("Observation predicted\nin {quant[1]}%-{quant[2]}% interval?"),
    fill = "Quantile interval"  
  )
  + ggplot2::scale_colour_discrete(labels=c(`FALSE`="Yes", `TRUE`="No"))
  + ggplot2::theme_classic(base_size=14)
)
```

## Other plotting functions

Behind the scenes, `plot_forecast()` relies on several helper functions to plot the different aspects of a forecast.
If you want further control over the visualization process, you can use these functions directly
They are as follows:^[We have omitted some optional paremeters below for brevity. See the function documentation for details.]

- `plot_ensemble(plt=NULL, fcst)` plots an ensemble of forecast realizations^[If `fcst` contains a `sim` column, the realizations will be plotted as individual curves. Otherwise, the forecast data will be plotted as points.]
- `plot_observations(plt=NULL, obs)` plots a set of observations. If those observations contain a `score` column, then the scores will be used to colour-code the observations
- `plot_quantiles(plt=NULL, fcst, quants=NULL)` plots the specified quantiles of the forecast. If `quants` is `NULL`, the quantiles will be inferred from `fcst`'s columns
- `plot_mean(plt=NULL, fcst)` plots the mean of the forecast data
- `plot_quant_intervals(plt=NULL, fcst, quant_pairs=NULL)` uses `ggplot2::geom_ribbon()` to plot bands of quantile intervals. Its `quant_pairs` parameter behaves the same as the parameter with the same name in `plot_forecast()`

All these functions consume a `ggplot2::ggplot` object as their first parameter, and return a `ggplot2::ggplot` object as well.
This makes it possible to compose these functions in any order to produce the plot you want.

```{r other_plot_examples}
fc <- create_forecast(list(
  time=1:3,
  vals = list(c(3,6,11), c(5,8,15), c(6,7,13), c(7,8,14), c(3,7,17))
))

obs <- data.frame(
  time=1:3,
  val_obs=c(4,10,14)
)

NULL |> plot_ensemble(fc) |> plot_mean(fc) |> plot_quant_intervals(fc, c(25,75)) |> plot_observations(accuracy(fc, obs, quant_pairs=c(25,75), summarize=FALSE))
```